\bf Data Assimilation for Modeling and Calibrating of Disease-altered Potassium Channel Kinetics in Mouse Cardiomyocytes

### 2022-09-28 writing/draft.tex
Abstract shortened

\begin{abstract}
Activities of cardiac voltage-gated potassium channels (K\textsubscript{v}) are essential in the electrical conduction system of the heart. Even slight changes in K\textsubscript{v} functions can cause fatal heart disease, such as long QT syndrome. Our recent studies found that altered glycosylation causes aberrant electrical signaling in mouse cardiomyocytes and leads to heart failure and early death. Activities of K\textsubscript{v} channels were also affected by altered glycosylation, and K\textsuperscript{+} currents were significantly reduced. Among major cardiac voltage-gated ion channels (VGIC), K\textsubscript{v} channels have a distinctive feature: various isoforms generate their own unique currents contributing to different phases of repolarization of the cardiac action potential (AP). On the other hand, the dominant isoform is responsible for Na\textsuperscript{+} and Ca\textsuperscript{2+} currents. This characteristic raises challenges for studying K\textsubscript{v} channels. Although it is highly desirable to observe the activities of each K\textsubscript{v} isoform separately, only collective outputs can be measured because of their slightly overlapping voltage-dependence of gating. In whole-cell recordings, the sum of K\textsuperscript{+} currents (I\textsubscript{Ksum}) can be measured. Traditional methods apply curve fitting to estimate components of K\textsubscript{v} isoforms from I\textsubscript{Ksum}. Additional curve fitting is used to find kinetics information for the individual K\textsuperscript{+} currents. However, this traditional approach is limited in its ability to determine gating kinetics of K\textsubscript{v} isoforms rigorously. Curve fittings do not account for interactions of multiple gating processes of various isoforms and dynamics according to changes in protocols. Therefore, we propose a novel data assimilation framework for investigating whole-cell K\textsubscript{v} current traces and apply the proposed method to our glycosylation data. First, computer models of K\textsuperscript{+} currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. Second, we develop fractional factorial designs to identify the parameters that have significant impacts on the current shapes. Third, we develop model calibration as nonlinear optimization that couple the \textit{in-silico} K\textsubscript{v} models and \textit{in-vitro} I\textsubscript{K} recordings. Last, low-dimensional embedding is used to visualize distributional difference of kinetic parameters of cells, showing the cellular-level variability. Experimental results show that the data assimilation method can estimate differences in kinetic rate between the control and experiment groups. This method has strong potential to pave a new way to analyze K\textsubscript{v} and heart diseases.
\end{abstract}

>>>>

\begin{abstract}
Potassium channels (K\textsubscript{v}) are responsible for repolarizing the cardiac action potential. Different K\textsubscript{v} isoforms contribute to various phases of repolarization. In the mouse ventricle, repolarization roughly consists of two kinetic components: the rapidly inactivating current I\textsubscript{Kto} (K\textsubscript{v}4.2), and the slowly inactivating currents such as I\textsubscript{Kslow1} and I\textsubscript{Kslow2} (K\textsubscript{v}1.5 and K\textsubscript{v}2.1). A data assimilation method, combining experimental data with mathematical/computational models, is used to estimate information on K\textsubscript{v} isoforms, because only the sum of their activities can be measured in the form of currents (I\textsubscript{K}). Most existing studies separate K\textsuperscript{+} currents mathematically from I\textsubscript{K} by a curve-fitting decomposition procedure. However, this traditional approach has limitations to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting decomposition only relies on the shape of K\textsuperscript{+} currents without considering underlying kinetics and interactions of K\textsubscript{v} isoforms, and 2) it handles I\textsubscript{K} traces from the same cell one-by-one individually. There is an urgent need to develop new methods that simultaneously model the underlying kinetics of multiple I\textsubscript{K} traces for a better understanding of pathological mechanisms. We propose a novel \textit{concurrent data assimilation} framework that utilizes biophysics-based computer models of K\textsubscript{v} to delineate gating kinetics in the multiple observations concurrently to cope with cellular-level dynamics. We apply the proposed framework to glyco-altered I\textsubscript{K} recordings from our previous studies. Experimental results show that the proposed method can discriminate characteristics between the control and experimental groups. This proposed method has strong potential to pave a new way to analyze K\textsubscript{v} and heart diseases.
\end{abstract}
